### 2018-04-18 concepts.tex
ok

Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
  binary-search that distinguishes delta-debugging \emph{per se}; the
  differences are immaterial for our purposes.}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
Delta-debugging in the context of detecting nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls (it is one of these ten
calls), but very difficult in a test consisting of a hundred library
function calls.  This is no different than the common use of
delta-debugging.  However, in horizontal nondeterminism detection,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  The behavior is part of a more general (and, to our
knowledge, not previously investigate) issue:  using delta-debugging
to minimize a test with respect to a predicate that only holds with a
certain probability (the predicate itself is nondeterministic).

@@

@@

It is the combination of
a one-way bias on faulty evaluations (the consequences of a false
positive for the predicate are much greater than for a false negative)
and the huge number of experiments relative to error rate, that
produces bad results, akin to the magnification of effect sizes
in science due to publication bias.  The bias in delta-debugging tends
towards producing a reduced test where probabilities are much smaller
than demanded by a predicate, due to the pressure on delta-debugging
to produce shorter tests.  Shorter tests have smaller probabilities,
on average, due to two factors:  first, timing-induced nondeterminism
has a much smaller temporal space to operate in, and second, if some
test operation introduces a small probability of nondeterminism, and
only many repetitions of that operation make the probability large,
small tests obviously on average have fewer instances of the
problematic operation.

@@

A costly but effective solution is to make $N$ large in comparison to
the number of expected predicate evaluations performed during
delta-debugging.  If the error rate is low enough, the problem
disappears.  However, given the large number of evaluations performed,
this will tend to make delta-debugging extremely slow.  We propose
using a dynamic sampling approach, where $N$ is small, but if the
predicate evaluates to true, $M$ repeated true evaluations
(``replications'', where the first true evaluation is counted as 1 replication) are
required before the predicate returns ``true'' to the delta-debugging
algorithm.  A set of $M$ repeated false positives with
$\frac{N}{M}$ samples each is much less likely than a false positive
with $N$ samples; so long as we accept the resulting bias in favor of
false negatives, we can therefore produce a reduced test with a
desired $P(fail)$ much more cheaply:  the use of replications not only
means we only pay the full sampling price on rare occasions, but a
desired accuracy for $P$ can be obtained with a much smaller value $M
\times N$ than a non-dynamically-sampled $N$.  For example, if we the predicate
is $P(fail) \geq 0.5$, and the true probability for a candidate test
is 0.25, using $N=8$ will give a false positive rate of over 10\%.
Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
positive rate of about 3.7\%, yet requires almost 60\% fewer
executions of the test\footnote{The probability calculations for such comparisons
  are relatively simple, but in real testing are
  usually not very useful, since the true
  probabilitiy distributions of test behaviors vary widely and
  dynamically during the
  delta-debugging process, making exact calculation unwieldy, even if
  the changing probabilities were known; exact results would be based on fictional
  probabilities, so gathering experimental data during a trial
  delta-debugging run and tuning $N$ and $M$ to yield desired
  results is  more effective.}.

@@

>>>>

Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
  binary-search that distinguishes delta-debugging \emph{per se}; the
  differences are immaterial for our purposes.}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
Delta-debugging in the context of detecting nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls (it is one of these ten
calls), but very difficult in a test consisting of a hundred library
function calls.  This is no different than the common use of
delta-debugging.  However, in horizontal nondeterminism detection,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  The behavior is part of a more general (and, to our
knowledge, not previously investigate) issue:  using delta-debugging
to minimize a test with respect to a predicate that only holds with a
certain probability (the predicate itself is nondeterministic).  Note
that while most descriptions of delta-debugging discuss reducing a
test with respect to \emph{test failure} (hence the term
``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
as reducing the length of a test while maintaining that an arbitrary
predicate remains true of the shorter versions of the test.  The
traditional predicate is ``the test fails'' but other predicates,
especially those related to code-coverage, can also be useful
\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution here is to consider
what happens when the predicate to be evaluated is not deterministic.
``The test behaves flakily'' is a simple, important, example of such a predicate.

@@

\begin{figure*}
\centering 
\begin{subfigure}{0.67\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddmin}
\caption{Standard delta-debugging}
\label{fig:p1}
\end{subfigure}
\begin{subfigure}{0.67\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforcep}
\caption{P=0.5, 100 samples}
\label{fig:p2}
\end{subfigure}
\begin{subfigure}{0.67\columnwidth}
\centering
\includegraphics[width=\columnwidth]{lengthddminforceprep}
\caption{10 samples, 10 replications}
\label{fig:p3}
\end{subfigure}
\caption{Delta-debugging of the same test, with different ways of
  handling nondeterministic in the reduction predicate.}
\end{figure*}

@@

It is the combination of
a one-way bias on faulty evaluations (the consequences of a false
positive for the predicate are much greater than for a false negative)
and the huge number of experiments relative to error rate, that
produces bad results, akin to the magnification of effect sizes
in science due to publication bias.  The bias in delta-debugging tends
towards producing a reduced test where probabilities are much smaller
than demanded by a predicate, due to the pressure on delta-debugging
to produce shorter tests.  Shorter tests have smaller probabilities,
on average, due to two factors:  first, timing-induced nondeterminism
has a much smaller temporal space to operate in (the test may be over
before a ``time-bomb'' set by an operation goes off, for example), and second, if some
test operations introduce a small probability of nondeterminism, and
only many repetitions of those operations make the probability large,
small tests obviously, on average, have fewer instances of the
problematic operations.

@@

A costly but effective solution is to make $N$ large in comparison to
the number of expected predicate evaluations performed during
delta-debugging.  If the error rate is low enough, the problem
disappears.  However, given the large number of evaluations performed,
this will tend to make delta-debugging extremely slow.  We propose
using a dynamic sampling approach, where $N$ is small, but if the
predicate evaluates to true, $M$ repeated true evaluations
(``replications'', where the first true evaluation is counted as 1 replication) are
required before the predicate returns ``true'' to the delta-debugging
algorithm.  A set of $M$ repeated false positives with
$\frac{N}{M}$ samples each is much less likely than a false positive
with $N$ samples; so long as we accept the resulting bias in favor of
false negatives, we can therefore produce a reduced test with a
desired $P(fail)$ much more cheaply:  the use of replications not only
means we only pay the full sampling price on rare occasions, but a
desired accuracy for $P$ can be obtained with a much smaller value $M
\times N$ than a non-dynamically-sampled $N$.  For example, if we the predicate
is $P(fail) \geq 0.5$, and the true probability for a candidate test
is 0.25, using $N=8$ will give a false positive rate of over 10\%.
Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
positive rate of about 3.7\%, yet requires almost 60\% fewer
executions of the test.  The probability calculations for such comparisons
  are relatively simple, but in real testing are
  usually not very useful, since the true
  probabilitiy distributions of test behaviors vary widely and
  dynamically during the
  delta-debugging process, making exact calculation unwieldy, even if
  the changing probabilities were known; exact results would be based on fictional
  probabilities, so gathering experimental data during a trial
  delta-debugging run and tuning $N$ and $M$ to yield desired
  results is  more effective.

@@

To make the basic concepts, more clear, Figures
\ref{fig:p1}-\ref{fig:p3} graphically show the interplay of
delta-debugging and nondeterministic predicates for a simple example.
In the example, tests consist of a sequence of operations that behave
nondeterministically with (independent) probabilities of 0.01, 0.05,
and 0.10, respectively, storing the value of the operation one of five array locations.
There is also an operation that clears out an array location, making
it possible to store another (possibly nondeterministic) value in that
array slot.  An example test sequence would be:

\begin{code}
array[2] = op01();
array[3] = op05();
clear(array[2]);
array[2] = op10();
\end{code}

The probability that this test behaves nondeterministically is
0.84645, the result of the simple calculation $(1-0.01) \times
(1-0.05) \times (1-0.10)$.  Figure \ref{fig:p1} shows one run of
delta-debugging on a test of length 500.  The initial test is
esentially guaranteed to behave nondeterministically; delta-debugging
only checks that it \emph{can} behave nondeterministically, and so, in
the course of reducing the test length to a test with only two
operations, the delta-debugging algorithm also reduces the probability
of nondeterminism to only about 20\%.  If we use a predicate that
``forces'' the test to behave nondeterministically at least half the
time, by sampling the predicate value 100 times and only returning
true when at least 50 of the evaluations report true, we see the
behavior in Figure \ref{fig:p2}:  the final test is slightly longer,
but still falls well short of our target of exhibiting nondeterminism
50\% of the time.  Finally, Figure \ref{fig:p3} shows what happens if
we use the same target of 50\% nondeterminism, but use only 10
samples, with 10 replications ($N$ = 10, $M$ = 10):  the test is not
much longer than in Figure \ref{fig:p2}, but the probability of
nondeterminism is above our target value, close to 60\%.

### 2018-04-18 implementation.tex
ok

Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
important source of nondeterminism that arises (only) from change in
process.  ASLR scrambles the layout of memory of the process in which
an executable runs in order to make it harder to exploit memory-safety
vulnerabilities in code.  As a side effect, it
means that a test that, in some circumstances, causes a crash, may at
other times not fail at all.

Another example of process-based nondeterminism was discovered by numerous Python programs when Python
version 3.3 introduced automatic random salting of hashes (see Section
\ref{sec:pnondet} per-process,
in order to mitigate hash-based denial of service attacks
\cite{denial}.  Until version 3.6 this not only resulted in changes in
exact hash values, but in the order of iteration on
dictionaries\footnote{In version 3.6, the default dictionary
  implementation became an ordered dictionary with consistent
  iteration, though actual hashes remained nondeterministic.}.  While
relatively few programs relied on exact hash values, many implicitly
relied on them in that they only functioned correctly with a
predictable order for dictionary iteration across processess.  Testing
Python code for nondeterminism based on this hash seed requires
running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
break all existing dictionaries and other structures relying on
hashing.

Other, less common, effects tied to process also exist.  A few tests
may somehow depend on their actual PID (Process ID) or at least the
parent process' ID.  Process change increases the likelyhood that on a multicore
system a test will execut on a different CPU than in the first
execution, which can have many subtle effects, mostly related to
changed timing.

@@

\caption{TSTL Method Calls for Nondeterminism Detection}
\label{tab:methods}
\begin{tabular}{l|l}
%\hline
Name & Purpose \\
\hline
\hline
{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
  tries with delay $d$\\
\hline
{\tt stepNondeterministic} & as above, except checks all visible values
                       (compares state at every step)\\
\hline
{\tt processNondeterministic} & as above, except runs test provided in
                          subprocess and compares non-opaque output
                          values \\
\hline
{\tt forceP} & wraps a predicate for test reduction (e.g. {\tt
         nondeterministic}) with desired probability of holding,
         sample size,\\
&  and replications required \\
\hline
\hline

@@

We implemented our approach in the TSTL \cite{NFM15} system, a
language and tool for property-based testing of Python code.
Because TSTL supports differential testing \cite{tstlsttt}, horizontal
nondeterminism detection can technically be implemented simply by
declaring a system to be its own reference, using TSTL's notation for
differential testing.  However, such an approach requires some effort
on the part of the user to control which values are checked for
equivalence, and does not easily support injecting timing differences,
or using a new process, in checking the behavior.

We therefore instead made nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  Tables \ref{tab:methods}
and \ref{tab:options} show, respectively, the primary additions to the
TSTL Python API and the TSTL command line tools.  The TSTL test
generation tool is called {\tt rt}, since the default mode is a fast
pure Random Tester, and {\tt reduce} is TSTL's tool for
delta-debugging and test
normalization (more aggressive reduction that also tends to ease
debugging and test triage by making tests failing due to the same
fault more similar \cite{onetest}).  With these additions,
developers of TSTL-based testing tools, or developers using TSTL tools
to test code can easily add nondeterminism to the set of properties
checked.  First-class vertical nondeterminism checks are currently
limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL
compiler to emit a harness where every action that causes an expected
exception to be raised is repeated to ensure the same exception is
still raised.  Users can also express that any action should exhibit
vertical determinism by wrapping the action in a function that checks
for failure and calls the code again.

>>>>

Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
important source of nondeterminism that arises (only) from change in
process.  ASLR scrambles the layout of memory of the process in which
an executable runs in order to make it harder to exploit memory-safety
vulnerabilities in code.  As a side effect, it
means that a test that, in some circumstances, causes a crash, may at
other times not fail at all.  Even if no inputs produce a crash,
however, memory errors may produce variation in values that are
overwritten or arise from uninitialized memory, and thus be detectable
as nondeterminism.

Another example of process-based nondeterminism was discovered by numerous Python developers when Python
version 3.3 introduced automatic random salting of hashes on a
per-process basis (a new salt chosen for each Python process),
in order to mitigate hash-based denial of service attacks
\cite{denial}.  Until version 3.6 this not only resulted in changes in
exact hash values, but in the order of iteration on
dictionaries\footnote{In version 3.6, the default dictionary
  implementation became an ordered dictionary with consistent
  iteration, though actual hashes remained nondeterministic.}.  While
relatively few programs rely on exact hash values, many implicitly
relied on them in that they only functioned correctly with a
predictable order for dictionary iteration.  Testing
Python code for nondeterminism based on this hash seed requires
running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
break all existing dictionaries and other structures relying on
hashing.

Other, less common, effects tied to process also exist.  A few tests
may somehow depend on their actual PID (Process ID) or at least the
parent process' ID.  Process change increases the likelyhood that on a multicore
system a test will execute on a different CPU than in the first
execution, which can have many subtle effects, mostly related to
changed timing.

@@

\caption{TSTL Method Calls for Nondeterminism Detection}
\label{tab:methods}
\begin{tabular}{l|l}
%\hline
Name & Purpose \\
\hline
\hline
{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
  tries with delay $d$\\
\hline
{\tt stepNondeterministic} & as above, except checks all visible values
                       (compares state at every step)\\
\hline
{\tt processNondeterministic} & as above, except runs test provided in
                          subprocess and compares non-opaque output
                          values \\
\hline
{\tt P} & returns probability that a predicate holds for a test (with
          optional number of sample to use in estimate) \\
\hline
{\tt forceP} & wraps a predicate for test reduction (e.g. {\tt
         nondeterministic}) with desired probability of holding,
         sample size,\\
&  and replications required \\
\hline
\hline

@@

We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
language and tool for property-based testing of Python code. TSTL has
been used to detect (and usually fix) errors in a number of widely
used Python libraries, the Python implementation itself, the Solidity
compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
Because TSTL supports differential testing \cite{tstlsttt}, horizontal
nondeterminism detection can technically be implemented simply by
declaring a system to be its own reference, using TSTL's notation for
differential testing.  However, such an approach requires some effort
on the part of the user to control which values are checked for
equivalence, and does not easily support injecting timing differences,
or using a new process, in checking the behavior.

We therefore instead made nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  Tables \ref{tab:methods}
and \ref{tab:options} show, respectively, the primary additions to the
TSTL Python API and the TSTL command line tools.  The TSTL test
generation tool is called {\tt rt}, since the default mode is a fast
pure Random Tester, and {\tt reduce} is TSTL's tool for
delta-debugging and test
normalization (more aggressive reduction that also tends to ease
debugging and test triage by making tests failing due to the same
fault more similar \cite{onetest}).  With these additions,
developers of TSTL-based testing tools, or developers using TSTL tools
to test code can easily add nondeterminism to the set of properties
checked.

First-class vertical nondeterminism checks are currently
limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL
compiler to emit a harness where every action that causes an expected
exception to be raised is repeated to ensure the same exception is
still raised.  Users can also express that any action should exhibit
vertical determinism by wrapping the action in a function that checks
for failure and calls the code again.

### 2018-04-18 introduction.tex
ok

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity.  If executing ``the same test'' can produce
significantly different behavior each time it is run, the consequences
can be unfortunate.  Developers using a test exhibiting nondeterminism
to debug a system
face a challenge in reasoning about causality, in that an event
observed may not even take place the next time the test is run.
Automated test generation systems using test coverage results to drive
the search for interesting inputs may be mislead when a run executes
code only rarely.  And, most significantly, regression testing
effectiveness can be significantly reduced if tests fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is sometimes problematic for
developers, who will usually want to  assume that library code they
use is deterministic; it is often disastrous for large-scale or highly
automated testing.

>>>>

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity.  If executing ``the same test'' can produce
significantly different behavior each time it is run, the consequences
can be unfortunate.  Developers using a test exhibiting nondeterminism
to debug a system
face a challenge in reasoning about causality, in that an event
observed may not even take place the next time the test is run; this
pernicious phenomenon has been dubbed the dreaded ``Heisenbug'' \cite{Heisenbug}.
Automated test generation systems using test coverage results to drive
the search for interesting inputs may be mislead when a run executes
code only rarely.  And, most significantly, regression testing
effectiveness can be significantly reduced if tests fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is sometimes problematic for
developers, who will usually want to  assume that library code they
use is deterministic; it is frequently vexing in debugging efforts; it is often disastrous for large-scale highly
automated testing.
